---
title: "Exploratory Analysis - ELO Difference"
output: html_notebook
---

# ELO Difference

## Purpose

The purpose of this analysis is to explore the following questions:
  * How do the ELO scores relate to game outcomes?
  * How do the ELO scores relate to win margin differences?
  * How do the ELO win probability and spread calculations compare to actuals?
  
  
## Setup

```{r}
# Load packages
source("../code/packages.R")
library(ggplot2)

# Load data
df_elo_diff <- read_parquet(here("data", "prepared", "df_elo_diff.parquet"))
```

  

## ELO Score Distributions

First let's explore the basic distribution of ELO difference for the team that is the favorite.

```{r}
df_elo_diff %>%
  ggplot(aes(x = favorite_elo_diff)) +
  geom_histogram(alpha = 0.5, color = "black") +
  theme_light() +
  labs(
    title = "ELO Difference Distribution (Favorite Team)",
    x = "Favorite ELO Difference",
    y = "Count"
  )

```

## Win Percentages

Now, let's layer in the win percentage with each bin of the histogram.

```{r}
df_elo_diff %>%
  mutate(bin = cut_interval(favorite_elo_diff, n = 30)) %>%
  group_by(bin) %>%
  summarize(
    count = n(),
    win_pct = sum(favorite_point_diff > 0) / count
  ) %>%
  ggplot(aes(x = bin, y = count, fill = win_pct)) +
  geom_col(alpha = 0.85) +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(
    title = "ELO Difference Distribution (Favorite Team)",
    x = "Favorite ELO Difference",
    y = "Count",
    fill = "Win %"
  )
```

The plot above shows a clear relationship between the pre-game ELO advantage and the win probability.  But how does this compare the estimated win probablity from the ELO score/diff?

```{r}
df_elo_diff %>%
  mutate(bin = cut_interval(favorite_elo_diff, n = 50)) %>%
  group_by(bin) %>%
  summarize(
    count = n(),
    win_pct_pred = mean(1 / (10^(-favorite_elo_diff / 400) + 1)),
    win_pct_actual = sum(favorite_point_diff > 0) / count
  ) %>%
  ungroup() %>%
  mutate(error = win_pct_actual - win_pct_pred) %>%
  ggplot(aes(x = win_pct_pred, y = win_pct_actual, color = error)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", alpha = 0.5) +
  theme_light() +
  labs(
    title = "Actual vs. Predicted Win Probability by ELO Difference",
    x = "Win Probability (Predicted)",
    y = "Win Probability (Actual)"
  )
```
The plot above confirms that there is a clear relationship between the ELO-estimated win probabilities and the actual outcomes.

## Point Differentials

Next, let's review the raw point differentials as a function of ELO difference.

```{r}
df_elo_diff %>%
  ggplot(aes(x = favorite_elo_diff, y = favorite_point_diff)) +
  geom_point(alpha = 0.05) +
  geom_density_2d() +
  theme_light() +
  labs(
    title = "Point Difference by ELO Difference",
    x = "ELO Difference",
    y = "Point Difference"
  )
```

Finally, let's check how the ELO-based spread estimate compares to the actual point differential.

```{r}
df_elo_diff %>%
  mutate(point_diff_pred = favorite_elo_diff / 25) %>%
  ggplot(aes(x = point_diff_pred, y = favorite_point_diff)) +
  geom_point(alpha = 0.05) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  labs(
    title = "Estimated Spread vs. Actual Point Differential",
    x = "Predicted Point Differential",
    y = "Actual Point Differential"
  )
```

